# Genome-wide methylation analysis report
- study: `r study`
- author: `r author`
- date: `r format(Sys.time(), '%d %B, %Y')`

## Parameters

```{r, results="markup"}
object$parameters
```

```{r}
out <- NULL
if (!is.null(object$sample.characteristics)) {
    out <- knit_child(file.path(report.path, "sample-characteristics.rmd"))
}
```

```{r,results="asis"}
if (!is.null(out))
    cat(out, sep="\n\n")
```


```{r}
out <- NULL
if (!is.null(object$covariate.associations)) {
    out <- knit_child(file.path(report.path, "covariate-associations.rmd"))
}
```

```{r,results="asis"}
if (!is.null(out))
    cat(out, sep="\n\n")
```


## QQ plots

```{r, include=FALSE}
plot <- object$qq.plot
out <- knit_child(file.path(report.path, "qq-plot.rmd"))
```

```{r,results="asis"}
cat(out, sep="\n")
```

## Manhattan plots

```{r, include=FALSE}
plot <- object$manhattan.plot
out <- knit_child(file.path(report.path, "manhattan-plot.rmd"))
```

```{r,results="asis"}
cat(out, sep="\n")
```

## Significant CpG sites

There were `r length(object$significant.sites)`
CpG sites with association p-values < `r object$parameters$sig.threshold`.
These are listed in the file [associations.csv](associations.csv).

```{r}
tab <- with(object, cpg.stats[match(significant.sites, rownames(cpg.stats)),])
write.csv(tab, file=file.path(opts_knit$get("output.dir"), "associations.csv"))
```

```{r}
practical.sites <- object$practical.sites
tab <- with(object, cpg.stats[match(practical.sites, rownames(cpg.stats)),])
```

Below are the `r length(practical.sites)`
CpG sites with association p-values < `r object$parameters$practical.threshold`
in the `r object$parameters$model` regression model.

```{r}
knitr::kable(tab)
```

Plots of these sites follow, one for each covariate set.
"p[lm]" denotes the p-value obtained using a linear model
and "p[beta]" the p-value obtained using beta regression.

```{r, include=FALSE}
out <- NULL
for (cpg in practical.sites) 
    out <- c(out, knit_child(file.path(report.path, "cpg-plot.rmd")))
```

```{r}
cat(out, sep="\n")
```

## Selected CpG sites

Number of CpG sites selected: `r length(object$selected.sites)`.

```{r}
tab <- with(object, cpg.stats[match(selected.sites, rownames(cpg.stats)),])
knitr::kable(tab)
```

```{r, include=FALSE}
out <- NULL
for (cpg in object$selected.sites)
    out <- c(out, knit_child(file.path(report.path, "cpg-plot.rmd")))
```

```{r}
cat(out, sep="\n\n")
```

## R session information

```{r, results="markup"}
sessionInfo()
```
